Demography and selection since maize domestication

### 2015-05-22 manuscript.tex
Singleton/pi figs updated

\subsection{Patterns of purifying and background selection are
  influenced by demographic history}
Purifying selection refers to the situation where deleterious
mutations arising in a population are continuously selected against. When this
form of selection is operating, it can serve to reduce genetic
variability at linked neutral sites, a phenomenon called
background selection \cite{charlesworth1993}. Purifying and background selection lead
to lower diversity within genes and other functional sites relative
to neutral regions \textcolor{red}{(ref here)}. We investigated purifying
selection in maize and teosinte by evaluating the average magnitude of reduced
diversity within genes and recovery away from genes in both taxa. When standardized by neutral
levels, which were defined as mean diversity at postions at least 0.01 cM distal
from genes, a stronger reduction of diversity
and slower recovery was observed for teosinte than for maize, implying
that purifying selection has left a more pronounced signature in the
teosinte genome (Figure \ref{purify}). This conflicted with our \emph{a priori} hypothesis;
we expected that strong artificial selection since domestication may
have elevated the intensity of purifying selection for maize. We therefore
conducted a parallel analysis based on singleton diversity. As a class,
singleton alleles depict the most recent patterns of evolution, but
also have the lowest effect on pairwise diversity. Therefore, unlike
pairwise diversity, patterns of singleton
diversity reflect recent patterns of evolution. Maize was downsampled
to reflect the expected level of singletons from a sample size of 13,
equivalent to that of teosinte, and for singletons our definition of
neutral diversity was modified to be mean diversity at positions at
least 0.02 cM distal from genes. When
evaluating the data in this manner, a very different pattern
emerged. Maize singleton diversity was just as low as teosinte singleton
diversity near genes, but recovered more slowly
(Figure \ref{purify}), implying that in the
recent past maize has been at least as influenced by purifying selection as
teosinte. Together, these findings imply that demographic history has
a strong influence on the effect of purifying selection. Historically,
teosinte has had a larger population size than maize, and only
recently has maize population size overcome that of teosinte. Since
the efficacy of purifying selection scales with population size,
these results likely reflect changes in $N_e$ more than they reflect
underlying changes in selection pressure.

\begin{figure*}[b]
%\vspace*{.05in}
\centering
\includegraphics[width=.45\textwidth]{FigsAndFiles/distanceToGene_WithSignificance_Folded2_manuscript.png} \includegraphics[width=.45\textwidth]{FigsAndFiles/distanceToGene_WithSignificance_Singletons_Downsampled_manuscript.png}
\caption{Relative level of diversity versus distance to the nearest
  gene, in maize and teosinte. Two measures of diversity were
  investigated. \textbf{A} displays pairwise
  diversity, which is most influenced by intermediate frequency
  alleles and therefore depicts more ancient evolutionary patterns,
  and \textbf{B} depicts singleton diversity, influenced by rare
  alleles and thus depicting evolutionary patterns in the recent past. \jri{are these done with latest values we used for curve fitting? should also add the singleton vs pi comparison within species as a supplemental figure}}
\label{purify}
\end{figure*}

@@

\subsection{Evaluating diversity around genes}
Two types of diversity surrounding genes were investigated. The first
was pairwise diversity in 1kb windows, as described previously. The
second was singleton diversity in 1kb windows. Singletons represent
the rarest class of alleles that this dataset can identify, and
collectively demonstrate the most recent patterns of evolution. Minor
allele frequencies were estimated with ANGSD \cite{korneliussen2014} using the
same quality filters previously described. Then, the number of
singletons in each non-overlapping 1kb window was calculated with R
\cite{R2014}. For maize, we generated a parallel set of downsampled
singleton data, for which binomial sampling within R was conducted
based on allele frequencies at each site, to generate a singleton
psuedo dataset of sample size 13, equivalent to our sample size for
teosinte. BiomaRt \cite{durinck2009, durinck2005} was then used to identify
the center of each gene. Next, the distance from each diversity window center
to the nearest gene center was computed.

>>>>

\subsection{Patterns of purifying and background selection are
  influenced by demographic history}
Purifying selection refers to the situation where deleterious
mutations arising in a population are continuously selected against. When this
form of selection is operating, it can serve to reduce genetic
variability at linked neutral sites, a phenomenon called
background selection \cite{charlesworth1993}. Purifying and background selection lead
to lower diversity within genes and other functional sites relative
to neutral regions \textcolor{red}{(ref here)}. We investigated purifying
selection in maize and teosinte by evaluating the average magnitude of reduced
diversity within genes and recovery away from genes in both taxa. When standardized by neutral
levels, which were defined as mean diversity at postions at least 0.01 cM distal
from genes, a stronger reduction of diversity
and slower recovery was observed for teosinte than for maize, implying
that purifying selection has left a more pronounced signature in the
teosinte genome (Figure \ref{purify}). This conflicted with our \emph{a priori} hypothesis;
we expected that strong artificial selection since domestication may
have elevated the intensity of purifying selection for maize. We therefore
conducted a parallel analysis based on singleton diversity. As a class,
singleton alleles depict the most recent patterns of evolution, but
also have the lowest effect on pairwise diversity. Therefore, unlike
pairwise diversity, patterns of singleton
diversity reflect recent patterns of evolution. Since our sample size
for maize (n=23) was larger than teosite (n=13), maize singletons were
analyzed directly as well as downsampled to reflect the sample size of
teosinte. For singletons, our definition of
neutral diversity was modified to be mean diversity at positions at
least 0.02 cM distal from genes. When
evaluating the data in this manner, a very different pattern
emerged. Maize singleton diversity was just at least as low as teosinte singleton
diversity near genes, but recovered more slowly
(Figure \ref{purify}), implying that in the
recent past maize has been at least as influenced by purifying selection as
teosinte. Together, these findings suggest that demographic history has
a strong influence on the effect of purifying selection. Historically,
teosinte has had a larger population size than maize, and only
recently has maize population size overcome that of teosinte. Since
the efficacy of purifying selection scales with population size,
these results likely reflect changes in $N_e$ more than they reflect
underlying changes in selection pressure.

\begin{figure*}[b]
%\vspace*{.05in}
\centering
\includegraphics[width=.45\textwidth]{FigsAndFiles/distanceToGene_WithSignificance_Folded2_manuscript.png} \includegraphics[width=.45\textwidth]{FigsAndFiles/distanceToGene_WithSignificance_Singletons_Downsampled_threeLines_manuscript.png}
\caption{Relative level of diversity versus distance to the nearest
  gene, in maize and teosinte. Two measures of diversity were
  investigated. \textbf{A} displays pairwise
  diversity, which is most influenced by intermediate frequency
  alleles and therefore depicts more ancient evolutionary patterns,
  and \textbf{B} depicts singleton diversity, influenced by rare
  alleles and thus depicting evolutionary patterns in the recent past. \jri{are these done with latest values we used for curve fitting? should also add the singleton vs pi comparison within species as a supplemental figure}}
\label{purify}
\end{figure*}

@@

\subsection{Evaluating diversity around genes}
Two types of diversity surrounding genes were investigated. The first
was pairwise diversity in 1kb windows, as described previously. The
second was singleton diversity in 1kb windows. Singletons represent
the rarest class of alleles that this dataset can identify, and
collectively demonstrate the most recent patterns of evolution. Minor
allele frequencies were estimated with ANGSD \cite{korneliussen2014} using the
same quality filters previously described. Then, the number of
singletons in each non-overlapping 1kb window was calculated with R
for both maize and teosinte \cite{R2014}. For maize, we also generated a parallel set of downsampled
singleton data, for which binomial sampling within R was conducted
based on allele frequencies at each site, to generate a singleton
psuedo dataset of sample size 13, equivalent to our sample size for
teosinte. BiomaRt \cite{durinck2009, durinck2005} was then used to identify
the center of each gene. Next, the distance from each diversity window center
to the nearest gene center was computed.
